New cardinality estimation algorithms for HyperLogLog sketches

### 2016-12-11 paper/paper.tex
minor changes

Under the assumption of a uniform hash function, the probability that the register values $\boldsymbol{\symRegValVariate} = (\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg)$ of a HyperLogLog sketch with parameters $\symPrecision$ and $\symRegRange$ are equal to $\boldsymbol{\symRegVal}=(\symRegVal_1,\ldots,\symRegVal_\symNumReg)$ is given by the corresponding probability mass function
\begin{equation}
\label{equ:multinomialProbabilityMass}
\symProbabilityMass(\boldsymbol{\symRegVal}\vert\symCardinality)
=
\sum_{\symCardinality_1+\ldots+\symCardinality_\symNumReg = \symCardinality} \binom{\symCardinality}{\symCardinality_1,\ldots,\symCardinality_\symNumReg}
\frac{1}{\symNumReg^\symCardinality}\prod_{\symIndexI=1}^\symNumReg \symRegProbability(\symRegVal_\symIndexI\vert\symCardinality_\symIndexI)
\end{equation}
where $\symCardinality$ is the cardinality. The $\symCardinality$ distinct elements are distributed over all $\symNumReg$ registers according to a multinomial distribution with equal probabilities. $\symRegProbability(\symRegVal\vert\symCardinality)$ is the probability that the value of a register is equal to $\symRegVal$, after it was selected $\symCardinality$ times by the insertion algorithm
\begin{equation}
\symRegProbability(\symRegVal\vert\symCardinality) 
:=
\begin{cases}
1 & \symCardinality=0 \wedge \symRegVal = 0\\
0 & \symCardinality=0 \wedge 1\leq\symRegVal\leq\symRegRange+1\\
0 & \symCardinality\geq1 \wedge \symRegVal = 0\\
\left(1-\frac{1}{2^\symRegVal}\right)^\symCardinality - \left(1-\frac{1}{2^{\symRegVal-1}}\right)^\symCardinality & \symCardinality \geq 1 \wedge 1\leq\symRegVal\leq\symRegRange\\
1 - \left(1-\frac{1}{2^{\symRegRange}}\right)^\symCardinality & \symCardinality\geq 1 \wedge \symRegVal = \symRegRange +1
\end{cases}
\end{equation}

@@

\subsection{Poisson approximation}
\label{sec:poisson_approximation}
Due to the statistical dependence of the register values, the probability mass function \eqref{equ:multinomialProbabilityMass} is difficult for further analysis. Therefore, a Poisson model can be used \cite{Flajolet2007}, which assumes that the cardinality itself is distributed according to a Poisson distribution
\begin{equation}
\symCardinality \sim \text{Poisson}(\symPoissonRate).
\end{equation}
Under the Poisson model the distribution of the register values is
\begin{align}
\symProbabilityMass(\boldsymbol{\symRegVal}\vert\symPoissonRate) 
&= 
\sum_{\symCardinality=0}^\infty \symProbabilityMass(\boldsymbol{\symRegVal}\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
\label{equ:poissonization}
\\
&= 
\sum_{\symCardinality_1=0}^\infty
\cdots
\sum_{\symCardinality_\symNumReg=0}^\infty
\prod_{\symIndexI=1}^\symNumReg
\symRegProbability(\symRegVal_\symIndexI\vert\symCardinality_\symIndexI)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^{\symCardinality_\symIndexI}}{\symCardinality_\symIndexI!\symNumReg^{\symCardinality_\symIndexI}}
\nonumber\\
&= 
\prod_{\symIndexI=1}^\symNumReg \sum_{\symCardinality=0}^\infty\symRegProbability(\symRegVal_\symIndexI\vert\symCardinality)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
\nonumber\\
&= 
\prod_{\symRegVal=0}^{\symRegRange+1} \left(
\sum_{\symCardinality=0}^\infty
\symRegProbability(\symRegVal\vert\symCardinality)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
\right)^{\!\!\symCount_\symRegVal}
\nonumber\\
\label{equ:poisson_pmf}
&=
e^{-\symCount_0\frac{\symPoissonRate}{\symNumReg}}
\left(\prod_{\symRegVal=1}^{\symRegRange}\left(e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)\right)^{\!\symCount_\symRegVal}\right)
\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}\right)^{\!\symCount_{\symRegRange+1}}.
\end{align}
Here $\symCount_\symRegVal$ denotes the multiplicity of value $\symRegVal$  in the multiset $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace$.
The final factorization shows that under the Poisson model the register values $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$ are independent and identically distributed. The probability that a register has a value less than or equal to $\symRegVal$ for a given rate $\symPoissonRate$ is defined by
\begin{equation}
\label{equ:register_value_distribution}
\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate)
=
\begin{cases}
0 & \symRegVal < 0 \\
e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}} & 0\leq \symRegVal \leq \symRegRange \\
1 & \symRegVal > \symRegRange
\end{cases}.
\end{equation}

@@

Let $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$ be a sample taken from \eqref{equ:assumed_register_val_distribution}. For large sample sizes $\symNumReg\rightarrow \infty$ we have asymptotically
\begin{equation}
\symExpectation\!\left(
\frac{1}{2^{-\symRegValVariate_1}+\ldots+2^{-\symRegValVariate_\symNumReg}}
\right)
\underset{\symNumReg\rightarrow \infty}{=}
\frac{1}{\symExpectation
\!\left(2^{-\symRegValVariate_1}+\ldots+2^{-\symRegValVariate_\symNumReg}\right)}
=
\frac{1}{\symNumReg \symExpectation\!\left(2^{-\symRegValVariate}\right)}.
\end{equation}
Together with \eqref{equ:expectation_power_two} we obtain
\begin{equation}
\symPoissonRate
=
\symExpectation\!\left(
\frac{\symAlpha_\infty\,\symNumReg^2\,\symPowerSeriesFunc\!\left(\log_2\!\left( \symPoissonRate/\symNumReg\right)\right)}{2^{-\symRegValVariate_1}+\ldots+2^{-\symRegValVariate_\symNumReg}}
\right)
\quad
\text{for $\symNumReg\rightarrow \infty$}.
\end{equation}
Therefore, the asymptotic relative bias of 
\begin{equation}
\label{equ:rawestimator_2}
\symPoissonRateEstimate 
= 
\frac{\symAlpha_\infty\,\symNumReg^2}{2^{-\symRegValVariate_1}+\ldots+2^{-\symRegValVariate_\symNumReg}}
\end{equation}
is bounded by $\symEpsPowerSeriesFunc$, which makes this statistic a good estimator for the Poisson parameter. It also corresponds to the raw estimator \eqref{equ:raw_estimator}, if the Poisson parameter estimate is used as cardinality estimate (see \cref{sec:depoissonization}).

@@

\cref{fig:raw_corrected_estimation_error_12_20} shows the distribution of the relative error of the estimated cardinality using \cref{alg:corrected_raw_estimation} compared to the true cardinality for $\symPrecision=12$ and $\symRegRange=20$. As the mean shows, the error is unbiased over the entire cardinality range. The new approach is able to accurately estimate cardinalities up to 4 billions ($\approx 2^{\symPrecision+\symRegRange}$) which is about an order of magnitude larger than the operating range upper bound of the raw estimator (\cref{fig:raw_estimate}) or the original method (\cref{fig:original_estimate}).

@@

\begin{figure}
\centering
\includesvg[width=1\textwidth]{stdev_comparison}
\caption{Standard deviations of the relative error of different cardinality estimators for a HyperLogLog sketch with paramters $\symPrecision = 12$ and $\symRegRange=20$ over the true cardinality.}
\label{fig:stdev_comparison}
\end{figure}

@@

Using the probability mass function of the Poisson model \eqref{equ:poisson_pmf} the log-likelihood and its derivative are given by
\begin{equation}
\label{equ:log_likelihood_single}
\log \mathcal{\symLikelihood}(\symPoissonRate\vert\boldsymbol{\symRegValVariate}) = 
-\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal}+ 
\sum_{\symRegVal=1}^{\symRegRange}\symCountVariate_\symRegVal \log\!\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)
+
\symCountVariate_{\symRegRange+1} \log\!\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegRange}}}\right)
\end{equation}
and
\begin{equation}
\frac{d}{d\symPoissonRate}\log \mathcal{\symLikelihood}(\symPoissonRate\vert\boldsymbol{\symRegValVariate}) 
=
-\frac{1}{\symPoissonRate}\left(
\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}}
+\symCountVariate_{\symRegRange+1}\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}}
\right).
\end{equation}
As a consequence, the maximum likelihood estimate for the Poisson parameter is given by 
\begin{equation}
\symPoissonRateEstimate = \symNumReg\symXEstimate,
\end{equation}
if $\symXEstimate$ is the root of the function
\begin{equation}
\label{equ:funcdef}
\symFunc(\symX)
:=
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\frac{\frac{\symX}{2^\symRegVal}}{1-e^{\frac{\symX}{2^\symRegVal}}}
+\symCountVariate_{\symRegRange+1}\frac{\frac{\symX}{2^\symRegRange}}{1-e^{\frac{\symX}{2^\symRegRange}}}.
\end{equation}
This function can also be written as
\begin{equation}
\label{equ:func}
\symFunc(\symX)
:=
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\symHelper\!\left(\frac{\symX}{2^\symRegVal}\right)
+
\symCountVariate_{\symRegRange+1}\symHelper\!\left(\frac{\symX}{2^\symRegRange}\right)
-
\left(\symNumReg-\symCountVariate_0\right),
\end{equation}
where the function $\symHelper(\symX)$ is defined as
\begin{equation}
\label{equ:helper}
\symHelper(\symX):=1-\frac{\symX}{e^{\symX}-1}.
\end{equation}
$\symHelper(\symX)$ is strictly increasing and concave as can be seen in \cref{fig:helper_function}. For non-negative values $\symX$ the function ranges from $\symHelper(0)=0$ to $\symHelper(\symX\rightarrow \infty)=1$.
\begin{figure}
\centering
\includesvg[width=0.6\textwidth]{helper}
\caption{The function $\symHelper(\symX)$.}
\label{fig:helper_function}
\end{figure}
Since the function $\symFunc(x)$ is also strictly increasing, it is obvious that there exists a unique root $\symXEstimate$ for which $\symFunc(\symXEstimate)=0$. The function is non-positive at 0 since $\symFunc(0)=\symCountVariate_0-\symNumReg\leq 0$ and, in case $\symCountVariate_{\symRegRange+1}<\symNumReg$ which implies $\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}>0$, the function is at least linearly increasing. $\symCountVariate_{\symRegRange+1}=\symNumReg$ corresponds to the case with all registers equal to the maximum value $\symRegRange+1$, for which the maximum likelihood estimate would be infinite.

@@

\subsection{Inequalities for the maximum likelihood estimate}
In the following lower and upper bounds for $\symXEstimate$ are derived.
Applying Jensen's inequality on $\symHelper$ in \eqref{equ:func} gives an upper bound for $\symFunc(\symX)$:
\begin{equation}
\symFunc(\symX)
\leq
\symX
\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}
+
\left(\symNumReg-\symCountVariate_0\right)\cdot
\symHelper\!\left(
\symX\cdot
\frac{
\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
+
\frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}
}
{\symNumReg-\symCountVariate_0}\right)
-
\left(\symNumReg-\symCountVariate_0\right).
\end{equation}
The left-hand side is zero, if $\symXEstimate$ is inserted. Resolution for $\symXEstimate$ finally gives the lower bound
\begin{equation}
\label{equ:strong_lower_bound}
\symXEstimate\geq \frac{\symNumReg-\symCountVariate_0}{\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
+
\frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}
}
\log\!\left(
1
+
\frac{\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
+
\frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}
}
{
\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
}
\right).
\end{equation}
This bound can be weakened using $\log(1+x) \geq \frac{2x}{x+2}$ for $x\geq0$ which results in
\begin{equation}
\label{equ:weak_lower_bound}
\symXEstimate
\geq
\frac{\symNumReg-\symCountVariate_0}
{\symCountVariate_0+\frac{3}{2}\sum_{\symRegVal=1}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal} + \frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}}.
\end{equation}
Using the monotonicity of $\symHelper$, the lower bound
\begin{equation}
\symFunc(\symX)
\geq
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)
+
\symCountVariate_{\symRegRange+1}\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)
-
\left(\symNumReg-\symCountVariate_0\right)
\end{equation}
can be found, where $\symRegValVariate'_\text{max} := \min(\symRegValVariate_\text{max}, \symRegRange)$ and
$\symRegValVariate_\text{max} := \max\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$.
Again, inserting $\symXEstimate$ and transformation gives
\begin{equation}
\label{equ:strong_upper_bound}
\symXEstimate
\leq
2^{\symRegValVariate'_\text{max}}
\log\!\left(
1+
\frac{\symNumReg-\symCountVariate_0}
{
2^{\symRegValVariate'_\text{max}}
\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
}
\right)
\end{equation}
as upper bound which can be weakened using $\log(1+\symX)\leq \symX$ for $\symX\geq 0$
\begin{equation}
\label{equ:weak_upper_bound}
\symXEstimate
\leq
\frac{\symNumReg-\symCountVariate_0}
{\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCountVariate_\symRegVal}{2^\symRegVal}}.
\end{equation}
If the HyperLogLog sketch is in the intermediate range, where $\symCountVariate_0=\symCountVariate_{\symRegRange+1}=0$ the bounds \eqref{equ:weak_lower_bound} and \eqref{equ:weak_upper_bound} differ only by a constant factor from the raw estimator \eqref{equ:raw_estimator}. Hence, the maximum likelihood method leads directly to the harmonic mean that is used by the raw estimator.

>>>>

Under the assumption of a uniform hash function, the probability that the register values $\boldsymbol{\symRegValVariate} = (\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg)$ of a HyperLogLog sketch with parameters $\symPrecision$ and $\symRegRange$ are equal to $\boldsymbol{\symRegVal}=(\symRegVal_1,\ldots,\symRegVal_\symNumReg)$ is given by the corresponding probability mass function
\begin{equation}
\label{equ:multinomialProbabilityMass}
\symProbabilityMass(\boldsymbol{\symRegVal}\vert\symCardinality)
=
\sum_{\symCardinality_1+\ldots+\symCardinality_\symNumReg = \symCardinality} \binom{\symCardinality}{\symCardinality_1,\ldots,\symCardinality_\symNumReg}
\frac{1}{\symNumReg^\symCardinality}\prod_{\symIndexI=1}^\symNumReg \symRegProbability(\symRegVal_\symIndexI\vert\symCardinality_\symIndexI)
\end{equation}
where $\symCardinality$ is the cardinality. The $\symCardinality$ distinct elements are distributed over all $\symNumReg$ registers according to a multinomial distribution with equal probabilities. $\symRegProbability(\symRegVal\vert\symCardinality)$ is the probability that the value of a register is equal to $\symRegVal$, after it was selected $\symCardinality$ times by the insertion algorithm
\begin{equation}
\symRegProbability(\symRegVal\vert\symCardinality) 
:=
\begin{cases}
1 & \symCardinality=0 \wedge \symRegVal = 0\\
0 & \symCardinality=0 \wedge 1\leq\symRegVal\leq\symRegRange+1\\
0 & \symCardinality\geq1 \wedge \symRegVal = 0\\
\left(1-\frac{1}{2^\symRegVal}\right)^\symCardinality - \left(1-\frac{1}{2^{\symRegVal-1}}\right)^\symCardinality & \symCardinality \geq 1 \wedge 1\leq\symRegVal\leq\symRegRange\\
1 - \left(1-\frac{1}{2^{\symRegRange}}\right)^\symCardinality & \symCardinality\geq 1 \wedge \symRegVal = \symRegRange +1.
\end{cases}
\end{equation}

@@

\subsection{Poisson approximation}
\label{sec:poisson_approximation}
Due to the statistical dependence of the register values, the probability mass function \eqref{equ:multinomialProbabilityMass} is difficult for further analysis. Therefore, a Poisson model can be used \cite{Flajolet2007}, which assumes that the cardinality itself is distributed according to a Poisson distribution
\begin{equation}
\symCardinality \sim \text{Poisson}(\symPoissonRate).
\end{equation}
Under the Poisson model the distribution of the register values is
\begin{align}
\symProbabilityMass(\boldsymbol{\symRegVal}\vert\symPoissonRate) 
&= 
\sum_{\symCardinality=0}^\infty \symProbabilityMass(\boldsymbol{\symRegVal}\vert\symCardinality) e^{-\symPoissonRate}\frac{\symPoissonRate^\symCardinality}{\symCardinality!}
\label{equ:poissonization}
\\
&= 
\sum_{\symCardinality_1=0}^\infty
\cdots
\sum_{\symCardinality_\symNumReg=0}^\infty
\prod_{\symIndexI=1}^\symNumReg
\symRegProbability(\symRegVal_\symIndexI\vert\symCardinality_\symIndexI)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^{\symCardinality_\symIndexI}}{\symCardinality_\symIndexI!\symNumReg^{\symCardinality_\symIndexI}}
\nonumber\\
&= 
\prod_{\symIndexI=1}^\symNumReg \sum_{\symCardinality=0}^\infty\symRegProbability(\symRegVal_\symIndexI\vert\symCardinality)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
\nonumber\\
&= 
\prod_{\symRegVal=0}^{\symRegRange+1} \left(
\sum_{\symCardinality=0}^\infty
\symRegProbability(\symRegVal\vert\symCardinality)e^{-\frac{\symPoissonRate}{\symNumReg}}\frac{\symPoissonRate^\symCardinality}{\symCardinality!\symNumReg^\symCardinality}
\right)^{\!\!\symCount_\symRegVal}
\nonumber\\
\label{equ:poisson_pmf}
&=
e^{-\symCount_0\frac{\symPoissonRate}{\symNumReg}}
\left(\prod_{\symRegVal=1}^{\symRegRange}\left(e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)\right)^{\!\symCount_\symRegVal}\right)
\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}\right)^{\!\symCount_{\symRegRange+1}}.
\end{align}
Here $\symCount_\symRegVal$ denotes the multiplicity of value $\symRegVal$  in the multiset $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace$.
The final factorization shows that under the Poisson model the register values $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$ are independent and identically distributed. The probability that a register has a value less than or equal to $\symRegVal$ for a given rate $\symPoissonRate$ is defined by
\begin{equation}
\label{equ:register_value_distribution}
\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate)
=
\begin{cases}
0 & \symRegVal < 0 \\
e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}} & 0\leq \symRegVal \leq \symRegRange \\
1 & \symRegVal > \symRegRange.
\end{cases}
\end{equation}

@@

Let $\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg$ be a sample distributed according to \eqref{equ:assumed_register_val_distribution}. For large sample sizes $\symNumReg\rightarrow \infty$ we have asymptotically
\begin{equation}
\symExpectation\!\left(
\frac{1}{2^{-\symRegValVariate_1}+\ldots+2^{-\symRegValVariate_\symNumReg}}
\right)
\underset{\symNumReg\rightarrow \infty}{=}
\frac{1}{\symExpectation
\!\left(2^{-\symRegValVariate_1}+\ldots+2^{-\symRegValVariate_\symNumReg}\right)}
=
\frac{1}{\symNumReg \symExpectation\!\left(2^{-\symRegValVariate}\right)}.
\end{equation}
Together with \eqref{equ:expectation_power_two} we obtain
\begin{equation}
\symPoissonRate
=
\symExpectation\!\left(
\frac{\symAlpha_\infty\,\symNumReg^2\,\symPowerSeriesFunc\!\left(\log_2\!\left( \symPoissonRate/\symNumReg\right)\right)}{2^{-\symRegValVariate_1}+\ldots+2^{-\symRegValVariate_\symNumReg}}
\right)
\quad
\text{for $\symNumReg\rightarrow \infty$}.
\end{equation}
Therefore, the asymptotic relative bias of 
\begin{equation}
\label{equ:rawestimator_2}
\symPoissonRateEstimate 
= 
\frac{\symAlpha_\infty\,\symNumReg^2}{2^{-\symRegValVariate_1}+\ldots+2^{-\symRegValVariate_\symNumReg}}
\end{equation}
is bounded by $\symEpsPowerSeriesFunc$, which makes this statistic a good estimator for the Poisson parameter. It also corresponds to the raw estimator \eqref{equ:raw_estimator}, if the Poisson parameter estimate is used as cardinality estimate (see \cref{sec:depoissonization}).

@@

\cref{fig:raw_corrected_estimation_error_12_20} shows the distribution of the relative error of the estimated cardinality using \cref{alg:corrected_raw_estimation} compared to the true cardinality for $\symPrecision=12$ and $\symRegRange=20$. As the mean shows, the error is unbiased over the entire cardinality range. The new approach is able to accurately estimate cardinalities up to 4 billions ($\approx 2^{\symPrecision+\symRegRange}$) which is about an order of magnitude larger than the operating range upper bound of the raw estimator (\cref{fig:raw_estimate}).

@@

\begin{figure}
\centering
\includesvg[width=1\textwidth]{stdev_comparison}
\caption{Standard deviations of the relative error of different cardinality estimators over the true cardinality for a HyperLogLog sketch with paramters $\symPrecision = 12$ and $\symRegRange=20$ .}
\label{fig:stdev_comparison}
\end{figure}

@@

Using the probability mass function of the Poisson model \eqref{equ:poisson_pmf} the log-likelihood and its derivative are given by
\begin{equation}
\label{equ:log_likelihood_single}
\log \mathcal{\symLikelihood}(\symPoissonRate\vert\boldsymbol{\symRegValVariate}) = 
-\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal}+ 
\sum_{\symRegVal=1}^{\symRegRange}\symCountVariate_\symRegVal \log\!\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)
+
\symCountVariate_{\symRegRange+1} \log\!\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegRange}}}\right)
\end{equation}
and
\begin{equation}
\frac{d}{d\symPoissonRate}\log \mathcal{\symLikelihood}(\symPoissonRate\vert\boldsymbol{\symRegValVariate}) 
=
-\frac{1}{\symPoissonRate}\left(
\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}}
+\symCountVariate_{\symRegRange+1}\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}}
\right).
\end{equation}
As a consequence, the maximum likelihood estimate for the Poisson parameter is given by 
\begin{equation}
\symPoissonRateEstimate = \symNumReg\symXEstimate,
\end{equation}
if $\symXEstimate$ denotes the root of the function
\begin{equation}
\label{equ:funcdef}
\symFunc(\symX)
:=
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\frac{\frac{\symX}{2^\symRegVal}}{1-e^{\frac{\symX}{2^\symRegVal}}}
+\symCountVariate_{\symRegRange+1}\frac{\frac{\symX}{2^\symRegRange}}{1-e^{\frac{\symX}{2^\symRegRange}}}.
\end{equation}
This function can also be written as
\begin{equation}
\label{equ:func}
\symFunc(\symX)
:=
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\symHelper\!\left(\frac{\symX}{2^\symRegVal}\right)
+
\symCountVariate_{\symRegRange+1}\symHelper\!\left(\frac{\symX}{2^\symRegRange}\right)
-
\left(\symNumReg-\symCountVariate_0\right),
\end{equation}
where the function $\symHelper(\symX)$ is defined as
\begin{equation}
\label{equ:helper}
\symHelper(\symX):=1-\frac{\symX}{e^{\symX}-1}.
\end{equation}
$\symHelper(\symX)$ is strictly increasing and concave as can be seen in \cref{fig:helper_function}. For non-negative values $\symX$ the function ranges from $\symHelper(0)=0$ to $\symHelper(\symX\rightarrow \infty)=1$.
\begin{figure}
\centering
\includesvg[width=0.6\textwidth]{helper}
\caption{The function $\symHelper(\symX)$.}
\label{fig:helper_function}
\end{figure}
Since the function $\symFunc(x)$ is also strictly increasing, it is obvious that there exists a unique root $\symXEstimate$ for which $\symFunc(\symXEstimate)=0$. The function is non-positive at 0 since $\symFunc(0)=\symCountVariate_0-\symNumReg\leq 0$ and, in case $\symCountVariate_{\symRegRange+1}<\symNumReg$ which implies $\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}>0$, the function is at least linearly increasing. $\symCountVariate_{\symRegRange+1}=\symNumReg$ corresponds to the case with all registers equal to the maximum value $\symRegRange+1$, for which the maximum likelihood estimate would be positive infinite.

@@

\subsection{Inequalities for the maximum likelihood estimate}
In the following lower and upper bounds for $\symXEstimate$ are derived.
Applying Jensen's inequality on $\symHelper$ in \eqref{equ:func} gives an upper bound for $\symFunc(\symX)$:
\begin{equation}
\symFunc(\symX)
\leq
\symX
\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}
+
\left(\symNumReg-\symCountVariate_0\right)\cdot
\symHelper\!\left(
\symX\cdot
\frac{
\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
+
\frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}
}
{\symNumReg-\symCountVariate_0}\right)
-
\left(\symNumReg-\symCountVariate_0\right).
\end{equation}
The left-hand side is zero, if $\symXEstimate$ is inserted. Resolution for $\symXEstimate$ finally gives the lower bound
\begin{equation}
\label{equ:strong_lower_bound}
\symXEstimate\geq \frac{\symNumReg-\symCountVariate_0}{\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
+
\frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}
}
\log\!\left(
1
+
\frac{\sum_{\symRegVal=1}^{\symRegRange}
\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
+
\frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}
}
{
\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
}
\right).
\end{equation}
This bound can be weakened using $\log(1+x) \geq \frac{2x}{x+2}$ for $x\geq0$ which results in
\begin{equation}
\label{equ:weak_lower_bound}
\symXEstimate
\geq
\frac{\symNumReg-\symCountVariate_0}
{\symCountVariate_0+\frac{3}{2}\sum_{\symRegVal=1}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal} + \frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}}.
\end{equation}
Using the monotonicity of $\symHelper$, the lower bound
\begin{equation}
\symFunc(\symX)
\geq
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)
+
\symCountVariate_{\symRegRange+1}\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)
-
\left(\symNumReg-\symCountVariate_0\right)
\end{equation}
can be found, where $\symRegValVariate'_\text{max} := \min(\symRegValVariate_\text{max}, \symRegRange)$ and
$\symRegValVariate_\text{max} := \max\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$.
Again, inserting $\symXEstimate$ and transformation gives
\begin{equation}
\label{equ:strong_upper_bound}
\symXEstimate
\leq
2^{\symRegValVariate'_\text{max}}
\log\!\left(
1+
\frac{\symNumReg-\symCountVariate_0}
{
2^{\symRegValVariate'_\text{max}}
\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCountVariate_\symRegVal}{2^\symRegVal}
}
\right)
\end{equation}
as upper bound which can be weakened using $\log(1+\symX)\leq \symX$ for $\symX\geq 0$
\begin{equation}
\label{equ:weak_upper_bound}
\symXEstimate
\leq
\frac{\symNumReg-\symCountVariate_0}
{\sum_{\symRegVal=0}^{\symRegRange}
\frac{\symCountVariate_\symRegVal}{2^\symRegVal}}.
\end{equation}
If the HyperLogLog sketch is in the intermediate range, where $\symCountVariate_0=\symCountVariate_{\symRegRange+1}=0$ the bounds \eqref{equ:weak_lower_bound} and \eqref{equ:weak_upper_bound} differ only by a constant factor from the raw estimator \eqref{equ:raw_estimator}. Hence, consequent application of the maximum likelihood method leads directly to the harmonic mean that is used by the raw estimator.
